\documentclass[10pt,twocolumn]{article}

% ============================================================
% Packages
% ============================================================
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{url}

\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\title{TurboQuant in Practice: Implementing 3-Bit KV Cache Compression\\in a Production LLM Inference Engine}

\author{
AmesianX\\
Independent Researcher\\
\texttt{amesianx@gmail.com}\\
\url{https://powerhacker.net}\\
\url{https://github.com/AmesianX/TurboQuant}
}

\date{April 2026}

% ============================================================
\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We present the first production-grade implementation of TurboQuant---a KV cache compression method based on Walsh-Hadamard Transform (WHT) and Lloyd-Max quantization---within llama.cpp, the most widely deployed open-source LLM inference engine. Our implementation compresses the KV cache to 3 bits per element, achieving \textbf{4.2$\times$ raw memory reduction} with only \textbf{6\% perplexity degradation} versus FP16 on Gemma~4 26B, and extends this to an effective \textbf{${\sim}4.74\times$} at inference time via hybrid rank-value compression. Beyond faithful implementation, we make four novel contributions: (1)~a \emph{dynamic attention sharpening} formula $\alpha(N) = 1 + c\sqrt{\ln N}$ derived from MMSE and Extreme Value Theory that adapts the softmax correction to the runtime context length; (2)~identification and resolution of a \emph{half-precision accumulation bug} in the flash attention kernel where FP16 arithmetic causes precision loss amplified by the IWHT butterfly into non-deterministic MoE expert routing; (3)~a \emph{CUDA-native TriAttention} integration that exploits Parseval preservation to compute trig-based importance scores directly on quantized, WHT-domain K blocks, eliminating the GPU$\to$CPU roundtrip that dominates prior reference implementations and cleanly separating the \emph{rank} axis (who to keep) from the \emph{value} axis (bits per element); and (4)~a complete engineering framework supporting 22 quantization type variants across five head dimensions (64--576), including double WHT per-head for $d{=}64$, asymmetric K/V configurations, and a hybrid AMX3\_1 block layout (50B WHT part + 58B polar part $=$ 108B) that carries both quantized values and the trig summaries needed for TriAttention scoring. We release the full implementation as open source at \url{https://github.com/AmesianX/TurboQuant}.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

Large Language Models (LLMs) with long context windows require substantial GPU memory for key-value (KV) caches. For a 26B-parameter Gemma~4 model with 262K context, the FP16 KV cache alone consumes 5.4~GiB---often exceeding the model weights themselves. TurboQuant~\cite{turboquant2026}, introduced at ICLR 2026 by Google DeepMind, proposes compressing KV cache entries to 3--4 bits using the Walsh-Hadamard Transform (WHT) followed by Lloyd-Max quantization, with an optional Quantized Johnson-Lindenstrauss (QJL) correction term.

While the theoretical framework is elegant, no production implementation existed prior to this work. Bridging the gap between a research paper and a shipping CUDA kernel requires solving numerous engineering challenges that the paper does not address: shared memory layout for WHT butterfly operations, precision requirements for the inverse transform, interaction with existing flash attention frameworks, and support for the diverse head dimensions found in modern architectures (64, 128, 256, 512, 576).

In this paper, we describe the complete implementation within llama.cpp's flash attention vector kernel (\texttt{fattn-vec}), report performance on real models, and present four contributions that go beyond the original paper:

\begin{enumerate}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \textbf{Dynamic Attention Sharpening} (\S\ref{sec:sharpening}): Quantization noise flattens the softmax distribution. We derive a context-length-adaptive correction $\alpha(N) = 1 + c\sqrt{\ln N}$ from Minimum Mean Square Error (MMSE) theory and Extreme Value Theory (EVT), eliminating the need for empirical tuning.

\item \textbf{Half-Precision Pathology} (\S\ref{sec:precision}): We identify that the upstream flash attention kernel's FP16 (\texttt{half2}) accumulation path causes precision loss that is \emph{amplified} by the IWHT butterfly into catastrophic non-determinism in Mixture-of-Experts routing. Forcing FP32 accumulation for TBQ value types resolves the issue and improves perplexity by 2\%.

\item \textbf{CUDA-native TriAttention on Quantized K} (\S\ref{sec:triattention}): We observe that Parseval's identity makes the trig-based importance scoring of TriAttention~\cite{triattention2026} invariant under our WHT rotation---the statistic can be read off the quantized K blocks directly. We design a hybrid 108-byte block (\texttt{AMX3\_1}) carrying both a 50-byte WHT-quantized key and a 58-byte polar summary, and implement the scoring, histogram-based Top-B threshold, and physical eviction entirely in CUDA. This eliminates the GPU$\to$CPU roundtrip that dominates reference implementations, and cleanly separates the \emph{rank} axis (who to keep) from the \emph{value} axis (bits per element).

\item \textbf{Multi-Variant Engineering} (\S\ref{sec:engineering}): We implement 22 TBQ/TBQP/AMX type variants across block sizes 64--576, including double WHT per-head for $d=64$, asymmetric K/V dimensions (GLM 576/512), per-block vs.\ global normalization, QJL residual correction, and the hybrid AMX block---all within the template-instantiated flash attention framework.
\end{enumerate}

% ============================================================
\section{Background}
\label{sec:background}

\subsection{TurboQuant}

TurboQuant~\cite{turboquant2026} compresses KV cache vectors through three steps:

\paragraph{1. Randomized WHT.} Given a $d$-dimensional vector $\mathbf{x}$, compute $\mathbf{w} = H_d \cdot (\mathbf{s} \odot \mathbf{x})$, where $H_d$ is the $d \times d$ Hadamard matrix and $\mathbf{s} \in \{-1, +1\}^d$ is a fixed random sign pattern. The WHT uniformizes the coefficient distribution regardless of the input structure, making all coordinates approximately i.i.d.\ Gaussian by the Lindeberg CLT.

\paragraph{2. Lloyd-Max Quantization.} Each WHT coefficient is quantized to $b$ bits using the optimal Lloyd-Max codebook for the Gaussian distribution. For $b=3$ (8 levels), the theoretical SQNR is 31.2; for $b=4$ (16 levels), 56.2.

\paragraph{3. QJL Correction (optional).} TurboQuant-prod (TBQP) uses $(b{-}1)$-bit Lloyd-Max quantization plus a 1-bit Quantized Johnson-Lindenstrauss projection on the residual, improving reconstruction MSE at the cost of additional attention score variance.

\paragraph{Decoding.} At attention time, the query is WHT-transformed to match the key domain: $\text{score} = \langle H_d(\mathbf{s} \odot \mathbf{q}),\ \text{dequant}(\mathbf{k}_\text{tbq}) \rangle / d$. For values, the weighted sum is computed in the WHT domain, then inverse-transformed: $\mathbf{v}_\text{out} = \mathbf{s} \odot H_d(\sum_k \alpha_k \cdot \text{dequant}(\mathbf{v}_k)) / d$.

\subsection{TriAttention}

TriAttention~\cite{triattention2026} observes that for trained transformers the attention weight a query $\mathbf{q}$ assigns to a key $\mathbf{k}$ is well approximated, up to a monotone function, by a small set of trigonometric moments of the pre-RoPE $\mathbf{q}$ and $\mathbf{k}$ components. By precomputing per-layer, per-head summaries $\phi_\ell(\mathbf{k})$ over a fixed set of frequencies $\{\omega_j\}$ and a running Z-normalization, a rank score $r_\ell(t)$ can be assigned to each token $t$ without re-running the attention itself. Tokens whose rank falls below a Top-$B$ threshold are evicted from the KV cache, yielding a roughly $2\times$ live-token compression on top of whatever per-element compression the cache already uses.

The published reference implementation performs this scoring on the host, reading the full-precision K cache back over PCIe on every trigger. In \S\ref{sec:triattention} we show that when the cache is already in the quantized WHT domain, the rank score is invariant under our sign-pattern + Hadamard rotation---it can be read off the blocks in place.

\subsection{Flash Attention Vector Kernel}

The llama.cpp \texttt{fattn-vec} kernel implements flash attention for single-token decoding (autoregressive generation). Each thread block processes one query head, iterating over KV pairs in chunks. Key design choices relevant to TBQ:

\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \texttt{V\_DOT2\_F32\_F16\_AVAILABLE}: When defined, VKQ accumulators use \texttt{half2} and shared memory uses \texttt{half}, halving register and shared memory usage. This is correct for FP16 KV but, as we show in \S\ref{sec:precision}, catastrophic for TBQ.
\item Template instantiation: Each (K-type, V-type) combination is a separate \texttt{.cu} file, allowing per-combination compile-time configuration.
\item Online softmax with running maximum tracking and rescaling.
\end{itemize}

% ============================================================
\section{System Design}
\label{sec:design}

\subsection{Query Preprocessing: GPU-Side WHT}

The original TurboQuant paper assumes WHT is performed during quantization (CPU-side). In a flash attention kernel, the \emph{query} must also be WHT-transformed at runtime. We implement this entirely in GPU registers and shared memory:

\paragraph{$d=256$ (128 threads, 2 elements each).}
\begin{enumerate}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item Sign multiply: $f_i \leftarrow q_i \cdot s_i$ (register)
\item Stage 7 (stride 128): register butterfly between $f_0, f_1$
\item Stages 0--4 (stride 1--16): \texttt{\_\_shfl\_xor\_sync} warp shuffle
\item Stages 5--6 (stride 32, 64): shared memory read/write with XOR addressing
\end{enumerate}

\paragraph{$d=512$ (128 threads, 4 elements each).}
Stages 8--7 (stride 256, 128) in registers, stages 0--4 via warp shuffle, stages 5--6 via shared memory. Total: 9 butterfly stages, zero global memory traffic.

The WHT result is stored directly into \texttt{Q\_reg} with the attention scale factor applied: $\texttt{Q\_reg}[i] = w_i \times \text{scale} / d$.

\subsection{K Scoring: WHT-Domain Dot Product}

For TBQ keys, the dot product is computed entirely in the WHT domain. The dequantized centroid values are multiplied against the pre-transformed query:
\begin{equation}
\text{score} = \sum_{i=0}^{d-1} c_{\text{idx}[i]} \cdot \text{norm} \cdot Q_{\text{wht}}[i]
\end{equation}
For TBQP keys, a second term is added from the QJL projection:
\begin{equation}
\text{score}_\text{tbqp} = \text{score}_\text{mse} + d_\text{qjl} \cdot \sum_{i} \text{sign}_\text{qjl}[i] \cdot Q_{\text{wht2}}[i]
\end{equation}
where $Q_{\text{wht2}}$ is a second WHT of the query with independent random signs, scaled by $\sqrt{\pi/2}$ (the QJL norm correction factor).

\subsection{V Accumulation and IWHT Decode}

Value vectors are dequantized and accumulated in the WHT domain using the standard flash attention online softmax. After the KV loop completes, the accumulated result undergoes inverse WHT:

\begin{equation}
\mathbf{v}_\text{out} = \frac{1}{d}\,\mathbf{s} \odot H_d\!\left(\frac{\sum_k e^{\alpha \cdot s_k} \cdot \mathbf{v}_k}{\sum_k e^{\alpha \cdot s_k}}\right)
\end{equation}

The IWHT uses the same butterfly structure as the forward transform (since $H_d$ is self-inverse up to normalization), followed by sign undo and division by $d$.

\subsection{Per-Block vs.\ Global Normalization}

For $d=512$ with TBQ3, each 256-element half of the WHT vector is normalized independently (\emph{per-block norm}), yielding two scale factors stored in \texttt{block\_tbq3\_0.d}. This is because the two halves can have significantly different energy after the 512-point WHT. For TBQP3, a \emph{global norm} is used instead, because the QJL residual correction operates across the full 512-element vector and requires consistent scaling.

% ============================================================
\section{Dynamic Attention Sharpening}
\label{sec:sharpening}

\subsection{The Softmax Flattening Problem}

Quantization noise in K cache entries adds independent noise $\epsilon_i$ to each attention score $s_i = \mathbf{q}^\top \mathbf{k}_i$. The observed scores become $\hat{s}_i = s_i + \epsilon_i$ where $\epsilon_i \sim \mathcal{N}(0, \sigma_\epsilon^2)$. The softmax distribution over $\hat{s}_i$ is flatter than over $s_i$: the noise raises the effective temperature, reducing the probability mass on the correct maximum.

\subsection{Derivation of $\alpha(N)$}

We seek a multiplicative correction $\alpha > 1$ such that $\text{softmax}(\alpha \hat{s}_i) \approx \text{softmax}(s_i)$.

\paragraph{MMSE Condition.} For the softmax to be unbiased, we need the signal-to-noise ratio of the sharpened scores to match the original:
\begin{equation}
\alpha = \frac{\text{Var}[s_i]}{\text{Var}[s_i] - \sigma_\epsilon^2} = \frac{1}{1 - 1/\text{SQNR}_\text{eff}}
\end{equation}
For large SQNR, $\alpha \approx 1 + 1/(2 \cdot \text{SQNR}_\text{eff})$ via first-order Taylor expansion.

\paragraph{EVT Correction.} The maximum noise among $N$ competing tokens grows as $\sqrt{2\ln N}$ by the Gumbel distribution. This increases the probability of a wrong token dominating the softmax. Incorporating this:
\begin{equation}
\alpha(N) = 1 + \frac{1}{2 \cdot \text{SQNR}_\text{eff}} \cdot \frac{\sqrt{\ln N}}{\sqrt{\ln N_0}}
\end{equation}

Defining $c = 1/(2 \cdot \text{SQNR}_\text{eff} \cdot \sqrt{\ln N_0})$ with reference $N_0 = 2048$:
\begin{equation}
\boxed{\alpha(N) = 1 + c \cdot \sqrt{\ln N}}
\label{eq:alpha}
\end{equation}

This is computed at runtime using the actual KV count \texttt{k\_VKQ\_max}:
\begin{verbatim}
float alpha = 1.0f + c * sqrtf(
    logf(fmaxf((float)k_VKQ_max, 2.0f)));
\end{verbatim}

\subsection{Per-Type Coefficients}

The coefficient $c$ depends on the effective SQNR, which differs between TBQ and TBQP due to the QJL noise source:

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Type & Structure & SQNR$_\text{eff}$ & $c$ & $\alpha$(2K) \\
\midrule
TBQ3  & 3-bit Lloyd-Max     & 31.2 & 0.00579 & 1.016 \\
TBQP3 & 2-bit + 1-bit QJL   & 13.8 & 0.01304 & 1.036 \\
TBQ4  & 4-bit Lloyd-Max     & 56.2 & 0.00326 & 1.009 \\
TBQP4 & 3-bit + 1-bit QJL   & 24.5 & 0.00724 & 1.020 \\
\bottomrule
\end{tabular}
\caption{Sharpening coefficients. TBQP types have lower effective SQNR because the QJL random projection adds a second noise source to the attention score.}
\label{tab:alpha}
\end{table}

\subsection{Properties}

The formula has several desirable properties:
\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \textbf{No clamp needed}: $\sqrt{\ln N}$ grows sub-logarithmically ($\sqrt{\ln 2} = 0.83$ to $\sqrt{\ln 262144} = 3.49$), keeping $\alpha$ in the narrow range 1.005--1.048.
\item \textbf{Autoregressive-safe}: During generation ($N$ small), $\alpha$ is small and conservative. During prefill ($N$ large), $\alpha$ increases to compensate the stronger noise effect.
\item \textbf{No hyperparameters}: All constants are derived from the quantization scheme (SQNR) and a reference context size ($N_0$).
\end{itemize}

% ============================================================
\section{Half-Precision Pathology in Flash Attention}
\label{sec:precision}

\subsection{The Bug}

The upstream \texttt{fattn-vec} kernel defines \texttt{V\_DOT2\_F32\_F16\_AVAILABLE} on CUDA architectures with native FP16 support. When active:
\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item VKQ accumulators are \texttt{half2} (16-bit)
\item Shared memory \texttt{KQ[]} is \texttt{half} (16-bit)
\item Query registers \texttt{Q\_reg} are \texttt{half2}
\end{itemize}

For standard FP16 KV caches, this is correct---the values are already in FP16, so no precision is lost. However, for TBQ values, three compounding precision losses occur:

\paragraph{Loss 1: VKQ Accumulation.} The weighted sum $\sum_k \text{softmax}_k \cdot \mathbf{v}_k$ is accumulated in \texttt{half2}. Each addition truncates to FP16 (10-bit mantissa), introducing rounding errors that accumulate over hundreds of KV entries.

\paragraph{Loss 2: Shared Memory Staging.} The reduce stage writes partial sums to \texttt{\_\_shared\_\_ half KQ[]}, then the IWHT reads them as \texttt{float*}---a type mismatch that further truncates values.

\paragraph{Loss 3: WHT Amplification.} The IWHT butterfly computes sums and differences of adjacent elements. In a $d$-point WHT, each output depends on \emph{all} $d$ inputs through $\log_2 d$ butterfly stages. A small rounding error $\epsilon$ in one input propagates to $O(\sqrt{d})$ error in the output. For $d=512$: $\sqrt{512} \approx 22.6\times$ amplification.

\subsection{Manifestation}

The amplified errors push MoE expert routing scores past decision boundaries, causing different experts to be selected across runs with identical inputs and \texttt{temp=0}. This produced \emph{completely different output sequences} (diverging at token positions as early as 23--27 out of ${\sim}50$), despite the kernel being mathematically deterministic.

\subsection{The Fix}

We force FP32 accumulation for all TBQ V type template instances by undefining the macro before the kernel header:
\begin{verbatim}
// In each TBQ V template instance .cu:
#include "../common.cuh"
#undef V_DOT2_F32_F16_AVAILABLE
#include "../fattn-vec.cuh"
\end{verbatim}

Due to \texttt{\#pragma once} on \texttt{common.cuh}, the re-include from \texttt{fattn-vec.cuh} does not re-define the macro. This selectively forces the FP32 path for TBQ V instances (70 files) while preserving the FP16 fast path for standard types.

Additionally, we implement float register staging for the IWHT input:
\begin{verbatim}
float tbq_regs[D_V / nthreads];
// reduce loop stores to tbq_regs
for (i0 = 0; i0 < D_V; i0 += nthreads)
    tbq_regs[i0/nthreads] = dst_val;
__syncthreads();
// write float values to shared memory
float *buf = (float*)&KQ[0];
for (i0 = 0; i0 < D_V; i0 += nthreads)
    buf[i0 + tid] = tbq_regs[i0/nthreads];
\end{verbatim}

\subsection{Impact}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
Configuration & PPL & vs FP16 \\
\midrule
FP16/FP16 baseline & 419.8 & 1.00$\times$ \\
half2 path (before fix) & 454.7 & 1.08$\times$ \\
\textbf{float path (after fix)} & \textbf{445.7} & \textbf{1.06$\times$} \\
\bottomrule
\end{tabular}
\caption{Perplexity improvement from forcing FP32 accumulation.}
\label{tab:precision}
\end{table}

Non-determinism was completely eliminated: FP16 baseline produces identical output across 20 runs, and TBQ now matches this behavior (with minor borderline variation at 3-bit quantization boundaries).

% ============================================================
\section{TriAttention on Quantized K}
\label{sec:triattention}

This section describes our CUDA-native port of TriAttention and the hybrid AMX3\_1 block that carries both a TBQ-style quantized key and the polar summary needed to score it. The framing is deliberately separated from the earlier WHT sections: TBQ reduces \emph{bits per element} (value axis); TriAttention reduces the \emph{number of retained tokens} (rank axis). The two compose multiplicatively at inference time.

\subsection{Key Insight: Parseval-Invariant Rank Score}

The TriAttention score for token $t$ in layer $\ell$ reduces, after Z-normalization, to a sum of inner products
\begin{equation}
r_\ell(t) = \sum_{j=1}^{J} \big( \langle \boldsymbol{\phi}^c_j, \mathbf{k}_t \rangle^2 + \langle \boldsymbol{\phi}^s_j, \mathbf{k}_t \rangle^2 \big),
\end{equation}
where $\boldsymbol{\phi}^c_j, \boldsymbol{\phi}^s_j$ are fixed cosine/sine vectors for frequency $\omega_j$. Under our cache transform $\mathbf{k}_t \mapsto \tilde{\mathbf{k}}_t = H_d (\mathbf{s} \odot \mathbf{k}_t)$, Parseval's identity gives
\begin{equation}
\langle \boldsymbol{\phi}, \mathbf{k}_t \rangle = \tfrac{1}{d}\langle H_d (\mathbf{s} \odot \boldsymbol{\phi}),\ \tilde{\mathbf{k}}_t \rangle,
\end{equation}
so if we pre-rotate each $\boldsymbol{\phi}$ by the same sign-pattern-then-Hadamard transform, the score is invariant. Crucially, the rotated $\tilde{\boldsymbol{\phi}}$ can be computed once at load time; the hot path needs only one WHT-domain inner product with the already-quantized $\tilde{\mathbf{k}}_t$, matching exactly the dot product we already compute for attention. No dequantization is required.

\subsection{Hybrid AMX3\_1 Block Layout}

A plain TBQ3\_1 block (128-element WHT with 3-bit codes) is 50 bytes and is sufficient to carry the WHT-domain key itself. TriAttention scoring, however, also needs a compact \emph{polar} summary---the $(r, \phi)$-form coordinates used to evaluate $\langle \boldsymbol{\phi}^c_j, \mathbf{k}_t \rangle$ and $\langle \boldsymbol{\phi}^s_j, \mathbf{k}_t \rangle$ without reconstructing the full vector.

We define a new block type \texttt{AMX3\_1} (head\_dim $=128$, 108 bytes total):
\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \textbf{Part~A (50 B)}: standard TBQ3\_1 payload---WHT sign/Hadamard, 3-bit Lloyd-Max indices, one norm per 128-element block. Used by the flash attention dot product exactly as before.
\item \textbf{Part~B (58 B)}: 3-bit magnitude $r$ per pair + 4-bit polar angle $\phi$ per pair + a 1-bit QJL correction. Used only by the TriAttention scoring kernel.
\end{itemize}

Part B is read only by the scoring kernel (triggered periodically, not per token), and Part A is read only by the attention kernel. The two parts share a common scale factor so that Part B's $r$ reproduces the same magnitudes the attention kernel would derive from Part A; this is what makes the hybrid internally consistent.

\subsection{CUDA-Native Trigger Path}

The scoring trigger runs every $\Delta$ tokens (default 256) and proceeds in three GPU passes over the cache, with no host roundtrip:

\begin{enumerate}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \textbf{Raw score + Z-normalize.} For each layer and each frequency $\omega_j$, a 1-D kernel reads Part~B of each block, evaluates $\langle \tilde{\boldsymbol{\phi}}^c_j, \tilde{\mathbf{k}}_t \rangle^2 + \langle \tilde{\boldsymbol{\phi}}^s_j, \tilde{\mathbf{k}}_t \rangle^2$, running-mean/variance over a window, and stores the layer score.
\item \textbf{Aggregate.} A reduce kernel sums the per-layer scores into a single rank $r(t)$ per token.
\item \textbf{Top-$B$ mask via histogram.} A 64-bin histogram of $r(t)$ is built on device; a prefix sum over the histogram yields the threshold $\tau$ such that $|\{t : r(t) \geq \tau\}| \approx B$. A mask kernel marks the remaining tokens for eviction.
\end{enumerate}

Only the threshold $\tau$ and the KV position array cross PCIe---a few KB per trigger, vs.\ the full-precision K cache (megabytes) in the host-based reference.

\subsection{Physical Eviction}

An evicted block must not be re-admitted on the next trigger. Simply zeroing the WHT payload (Part~A) is insufficient: the polar Part~B survives and can push the block above $\tau$ on the next pass. We zero both Part~A and Part~B (fields \texttt{d\_wht} and \texttt{d\_r}), making eviction permanent. This matters in practice: without it, the 95\% eviction ratio we observe at Top-$B = 4096$ drifts back toward 70\% within a few hundred new tokens.

\subsection{Attention Sink (\texttt{keep\_first})}

TriAttention alone is not enough for generative stability: pure Top-$B$ scoring will eventually evict the first few tokens, at which point the softmax collapses---empirically, Qwen3-14B produces "다름, 다름, 다름\ldots" repetition loops. This matches the StreamingLLM~\cite{streamingllm2023} finding that the earliest tokens act as attention sinks. We add an override kernel that forces the first $K_0$ positions to always be kept regardless of score; $K_0 = 32$ suffices for Qwen3-14B (4 is too small, 16 marginal, 32 fully stable).

\subsection{Why GPU-Native Matters}

Our measurements put the per-trigger cost at $\sim$2\% of decode latency at $B = 4096$, versus seconds of wall time for the reference host-based implementation at 14B scale. Three factors contribute:

\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \textbf{In-place block read.} The scoring kernel consumes the same quantized block bytes that are already resident in HBM for attention; no dequantization, no PCIe crossing.
\item \textbf{Stream-local mask $\to$ eviction.} The mask never leaves the GPU---the eviction kernel reads it directly from the same stream.
\item \textbf{Rank--value orthogonality.} Because Part~A is exactly TBQ3\_1, the existing flash attention kernel is unchanged by the addition of TriAttention; the two features compose rather than interfere.
\end{itemize}

We revisit the cost on the \emph{memory} side in \S\ref{sec:results}: AMX3\_1 is 108 B vs.\ TBQ3\_1's 50 B, so the raw bytes-per-token increases; the win comes from retaining fewer tokens.

% ============================================================
\section{Engineering Framework}
\label{sec:engineering}

\subsection{Type System}

We define 20 quantization types organized by block size:

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llcl@{}}
\toprule
Suffix & Block Size & Head Dim & Notes \\
\midrule
\_0 & 256 & 256, 512 & Primary type \\
\_1 & 128 & 128 & Standard transformers \\
\_2 & 64  & 64  & Small-head models \\
\_3 & 64  & 64  & Double WHT per-head \\
\_4 & 576 & 576 & GLM split 256+256+64 \\
\midrule
AMX3\_1  & 128 (108\,B) & 128 & Hybrid WHT+polar (K-side, v1.7.0) \\
AMXV3\_1 & 128 (50\,B)  & 128 & V-side AMX companion \\
\bottomrule
\end{tabular}
\caption{TBQ/AMX block variants. Each \_$i$ suffix has TBQ3/TBQ4/TBQP3/TBQP4 sub-types (3-bit, 4-bit, with/without QJL); 22 variants total. AMX3\_1 is the hybrid block used with TriAttention (\S\ref{sec:triattention}).}
\label{tab:types}
\end{table}

\subsection{Double WHT Per-Head (\_3 Types, $d=64$)}

\textbf{Note:} The cross-head WHT approach ($H_{512} = H_8 \otimes H_{64}$) described in earlier versions has been \textbf{abandoned} due to Q-K domain mismatch---Q operates in the 64-dim per-head space while K was in the 512-dim cross-head space. No IWHT reconstruction could bridge this gap.

\paragraph{Replacement: Double WHT per-head.} We apply two rounds of WHT with different sign patterns: $S_1 \to \text{WHT}_{64} \to S_2 \to \text{WHT}_{64}$. This achieves kurtosis $0.375 \to 0.047$ (near-Gaussian), much better decorrelation than single WHT. Both Q and K undergo the same double WHT, so Parseval's theorem applies directly---no inverse transform needed. Speed impact is negligible ($\sim$384 extra FLOPs per head).

\paragraph{QJL critical at $d=64$.} Contrary to earlier findings, QJL's 1-bit correction is \emph{essential} for multi-turn stability. Without QJL (TBQ3\_3), conversations collapse into repetition loops after 3--4 turns. With QJL (TBQP3\_3), 7+ turns verified with correct responses. K is automatically mapped to TBQP3\_3 at $d=64$.

\subsection{MMSE Dynamic Softening ($d=64$)}

At $d \geq 256$, quantization noise is small relative to the attention gap, and \emph{sharpening} ($\alpha > 1$) compensates softmax flattening. At $d=64$, the noise is large enough to corrupt the token ranking itself. Sharpening then amplifies confidence in wrong tokens, causing repetition loops.

We derive an MMSE-optimal \emph{softening} correction:
\begin{equation}
\alpha(N) = \frac{\text{SQNR}}{\text{SQNR} + \sqrt{\ln N / \ln N_0}}
\end{equation}
where $\text{SQNR}$ is the effective signal-to-quantization-noise ratio at $d=64$, and $N_0 = 2048$. This gives $\alpha < 1$: the softmax is made \emph{flatter}, reducing overconfidence. The EVT term $\sqrt{\ln N / \ln N_0}$ increases softening for longer contexts where more noise-boosted candidates compete.

This creates a unified framework: the same MMSE/EVT theory produces sharpening at high SQNR ($d \geq 256$) and softening at low SQNR ($d = 64$).

\subsection{Asymmetric K/V Dimensions}

GLM-4 uses $d_K=576, d_V=512$ with MLA (Multi-head Latent Attention). We support this via the \texttt{D\_V} template parameter, with the 576-dimensional K using a split WHT (256+256+64) and the 512-dimensional V using standard 512-point WHT.

\subsection{SWA F16 Bypass}

Gemma~4 uses Sliding Window Attention (SWA) for 25 of 35 layers. Despite occupying only 300~MiB (vs.\ 5120~MiB for global KV), SWA layers dominate output quality because they attend to the most recent, high-relevance tokens. We automatically upgrade SWA K/V to FP16, eliminating quantization error where it matters most.

\subsection{V Rotation Incompatibility}

The standard llama.cpp KV cache applies RoPE rotation to both K and V. For K, this is safe: RoPE is orthogonal, so quantizer inputs retain their norm and Gaussianization spectrum regardless of token position, and Parseval's identity preserves the relative-position inner product $\mathbf{q}_m^\top R_{n-m}\, \mathbf{k}$ across the WHT-domain dot product (the $\pm 1$ sign pattern $\mathbf{s}$ applied to both sides cancels, and the Hadamard matrix is orthogonal). For V, however, the IWHT decode has no corresponding inverse rotation to undo the RoPE that was applied on cache write, causing the output to be in a position-dependent rotated coordinate frame. We disable V rotation (\texttt{attn\_rot\_v=false}) for all TBQ configurations. The same argument explains why our internal WHT types (TBQ/TBQP/AMX) must bypass llama.cpp's external \texttt{attn\_rot\_k} path: applying two orthogonal rotations back-to-back destroys the per-coordinate centroid assumption of Lloyd-Max.

% ============================================================
\section{Development Timeline}
\label{sec:timeline}

The implementation evolved over eight days across five major sessions, each solving distinct engineering challenges:

\paragraph{v1.4.0 (Apr 1--3): Foundation + GLM 576.}
Initial CUDA kernel implementation for head\_dim=256 (Qwen, LLaMA) and 128 (standard transformers). GLM-4.7-Flash required head\_dim=576 support---not a power of two. We designed a split WHT: $576 = 256 + 256 + 64$, with independent sub-blocks for WHT and shared sign patterns. Simulation over 500 trials confirmed $-3.4\%$ MSE vs.\ full QR orthogonal rotation with zero memory overhead. During testing, discovered V cache IWHT was producing garbage due to FP16 shared memory precision---the first hint of what would become the central bug of v1.5.2.

\paragraph{v1.5.0 (Apr 4): Upstream Rebase + Gemma 4.}
Complete 3-way merge rebase onto upstream llama.cpp (commit \texttt{b7ad48ebd}), resolving 26 conflicts. Added Gemma~4 support: hybrid SWA architecture with global head\_dim=512 (10 layers) + SWA head\_dim=256 (50 layers), variable GQA (16/4 heads per layer). Key fixes: SWA cache type auto-remapping, variable GQA bypass for TBQ rotation, and 512-point single-pass WHT (9 butterfly stages, 128 threads $\times$ 4 elements).

\paragraph{v1.5.1 (Apr 4, PM): SWA F16 Bypass.}
Breakthrough: SWA cache occupies only 300~MiB (vs.\ 5120~MiB global) but dominates output quality through 25 layers of recent-token attention. Automatically upgrading SWA K/V to FP16 eliminated SWA quantization error entirely, achieving f16-equivalent math accuracy (37.4/65 vs.\ f16 36.6/65) for the first time.

\paragraph{v1.5.2 Session 1 (Apr 7): Attention Sharpening.}
PPL investigation revealed 21\% gap was not from bit depth but from softmax flattening. Sub-block scale experiment failed (WHT already uniformizes variance). Derived $\alpha = 1 + 1/(2 \times \text{SQNR})$ from MMSE theory. Discovered V rotation bug (\texttt{attn\_rot\_v}). PPL: 509.9 $\to$ 454.7.

\paragraph{v1.5.2 Session 2 (Apr 7--8): Precision Fix.}
Non-deterministic output with temp=0 led to exhaustive debugging: eliminated CUDA graphs, pool allocation, threadfence, MoE MUL\_MAT\_ID, SET\_ROWS, and dynamic sharpening as causes. Root cause: \texttt{half2} accumulation in upstream \texttt{fattn-vec} amplified by IWHT butterfly. Fixed with FP32 forced path for 70 TBQ V template instances. PPL: 454.7 $\to$ 445.7. Dynamic $\alpha(N)$ formula implemented.

\paragraph{v1.5.3 (Apr 11--12): Double WHT for $d=64$ + Side Buffer Experiment.}
Cross-head WHT ($H_{512} = H_8 \otimes H_{64}$) abandoned due to Q-K domain mismatch at $d=64$. Replaced with double WHT per-head: $S_1 \to \text{WHT}_{64} \to S_2 \to \text{WHT}_{64}$ (kurtosis 0.375 $\to$ 0.047). QJL re-enabled for K at $d=64$ (critical for multi-turn stability---9+ turns verified). Experimental f16 side buffer achieved 35/35 math (stored in branch \texttt{experiment/side-buffer-f16}). Key discovery: \texttt{src1} index mapping required for KV position consistency. Recommended production config for $d=64$: \texttt{--cache-type-k tbq4 --cache-type-v tbq3} (35/35 math verified on GPT-OSS 120B).

\paragraph{v1.6.0 (Apr 14): Polar Derotate Exploration---Anomalous Math Result.}
Before learning of TriAttention, we independently pursued the idea that RoPE-rotated K could be quantized more accurately if its phase were explicitly separated from its magnitude---\emph{polar derotate}. A prototype encoder split each K pair into $(r, \phi)$ form, quantized $r$ with Lloyd-Max and $\phi$ with a uniform angular grid, and explicitly undid the position rotation at dequantization. On a Qwen3-14B math benchmark (the same 35-problem suite), FP16 baseline scored $\sim$13/35, while our polar-derotate cache reached \textbf{31/35}---substantially higher than both TBQ3 and, surprisingly, the uncompressed FP16 baseline. We could not reproduce this advantage on other models or fully explain the mechanism (we suspect an implicit regularization from the angular grid, but this remains an open question; see \S\ref{sec:discussion}). After the TriAttention result in v1.7.0 provided a sounder rank-axis story, the dedicated polar-derotate type was retired; the AMX3\_1 block in v1.7.0 reuses the polar summary for a different purpose (TriAttention scoring).

\paragraph{v1.7.0 (Apr 20--21): TriAttention Integration + Hybrid AMX Block.}
After discovering the domvox reference implementation of TriAttention in Python, we ported the full trigger loop to CUDA (\S\ref{sec:triattention}), using Parseval's identity to score directly on quantized WHT blocks. A new hybrid block type \texttt{AMX3\_1} (108 B = 50 B TBQ3\_1 payload + 58 B polar summary) carries both the quantized key and the statistics the scoring kernel needs. On Qwen3-14B at 262K context, the effective compression reaches $\sim$2.37$\times$ raw $\times$ $\sim$2$\times$ live-token $\approx$ 4.74$\times$, with needle-in-haystack recall preserved once \texttt{keep\_first} $= 32$ is set to avoid sink-token eviction (4 was too small; 16 marginal; 32 fully stable). Housekeeping in the same release: TBQ/TBQP/AMX types now bypass the external \texttt{attn\_rot\_k} path (they do their own WHT internally; the external rotation was a redundant second orthogonal transform), and the defunct TBQX3\_1 scaffolding is removed.

% ============================================================
\section{Experimental Results}
\label{sec:results}

All experiments use NVIDIA DGX Spark (GB10, 128GB VRAM), CUDA 12.8, and Gemma~4 26B-A4B-it (MoE, UD-Q4\_K\_XL quantization) with 262K context.

\subsection{Perplexity}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
Configuration & PPL (wiki-2K) & vs FP16 \\
\midrule
FP16/FP16 & 419.8 & 1.00$\times$ \\
\midrule
v1.5.1 (before sharpening) & 509.9 & 1.21$\times$ \\
+ per-block norm & 498.2 & 1.19$\times$ \\
+ attention sharpening & 455.0 & 1.08$\times$ \\
+ V rotation fix & 448.1 & 1.07$\times$ \\
+ dynamic $\alpha$ & 454.7 & 1.08$\times$ \\
+ \textbf{FP32 accumulation} & \textbf{445.7} & \textbf{1.06$\times$} \\
\bottomrule
\end{tabular}
\caption{Progressive PPL improvement from v1.5.1 to v1.5.2.}
\label{tab:ppl_progress}
\end{table}

\subsection{Math Accuracy}

We evaluate on a 35-problem math benchmark (2$\times$2/3$\times$3 matrix multiplication, scalar arithmetic, with 0--2000 token filler) at temp=0:

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Config & Model & Avg (/35) & Peak & 10-run scores \\
\midrule
FP16 & Gemma 4 26B & 20.1 & 21 & 19--21 \\
\textbf{TBQP3/TBQ3} & Gemma 4 26B & \textbf{19.1} & \textbf{23} & 16--23 \\
\midrule
FP16 & GPT-OSS 120B & 35/35 & 35 & --- \\
\textbf{TBQ4/TBQ3} & GPT-OSS 120B & \textbf{35/35} & \textbf{35} & --- \\
TBQP3\_3/TBQ3 & GPT-OSS 120B & \multicolumn{3}{c}{Korean {\checkmark}, matrix math {\texttimes}} \\
\midrule
FP16                    & Qwen3-14B & $\sim$13 & 14 & 11--14 \\
\textbf{Polar-Derotate}$^\dagger$ & Qwen3-14B & --- & \textbf{31} & 28--31 \\
AMX3\_1 + TriAttention  & Qwen3-14B & --- & 13 & 11--14 \\
\bottomrule
\end{tabular}
\caption{Math benchmark results. On Gemma 4, TBQ peak (23) exceeds FP16 best (21). On GPT-OSS 120B ($d=64$), TBQ4 K achieves f16-equivalent 35/35. On Qwen3-14B, a prototype polar-derotate K cache ($^\dagger$v1.6.0, since retired) produced an \emph{unexplained} jump to 31/35---substantially above the FP16 baseline of $\sim$13/35. We do not yet have a theoretical account of this effect; we report it as an observation. The production v1.7.0 AMX3\_1 block matches FP16 on this bench while reducing memory via TriAttention.}
\label{tab:math}
\end{table}

\paragraph{Needle-in-haystack recall (Qwen3-14B, v1.7.0).}
We embed a 6-digit password in a 200K-token context and ask the model to retrieve it after eviction has occurred. With \texttt{keep\_first}$=4$ the model loses the sink tokens and degenerates into a repetition loop; with \texttt{keep\_first}$=16$ recall is flaky; with \texttt{keep\_first}$=32$ recall is stable across all 10 trials.

\subsection{Memory}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
Config & Global KV & Total KV & Compression \\
\midrule
FP16/FP16 & 5,120 MiB & 5,420 MiB & 1.0$\times$ \\
TBQP3/TBQ3 & 990 MiB & 1,290 MiB & 4.2$\times$ \\
\bottomrule
\end{tabular}
\caption{KV cache memory at 262K context (Gemma 4 26B MoE).}
\label{tab:memory}
\end{table}

\paragraph{Effective compression under TriAttention.}
On Qwen3-14B the AMX3\_1 block is 108 B vs.\ TBQ3\_1's 50 B, so the \emph{raw} cache is only $\sim$2.37$\times$ smaller than FP16---worse than TBQ3 alone. The win is on the \emph{live-token} axis: TriAttention evicts $\sim$50\% of slots once the context fills, giving an effective $\sim$2.37$\times$ $\times$ $\sim$2$\times$ $\approx$ 4.74$\times$ end-to-end compression vs.\ FP16. This is the framing we use in the abstract. The tradeoff is honest: we trade bytes-per-token for tokens-per-cache, and come out ahead because the rank axis compresses further than any remaining bit-level headroom.

\subsection{Trigger Overhead}

The TriAttention trigger on Qwen3-14B at $B=4096$ costs $\sim$2\% of wall-clock decode time per invocation (every 256 tokens). Decode throughput holds at $\sim$20.7 tok/s, within a few percent of TBQ3-only. The reference host-based implementation at the same scale stalls for seconds per trigger because of the GPU$\to$CPU K-cache roundtrip.

\subsection{Determinism}

With the FP32 accumulation fix, 20 consecutive runs with temp=0 produce identical output for all but 1--2 borderline tokens near 3-bit quantization boundaries. FP16 baseline achieves 20/20 identical runs. Prior to the fix, extreme divergence occurred as early as token 23.

% ============================================================
\section{Discussion: Open Questions}
\label{sec:discussion}

\paragraph{The polar-derotate math anomaly.} The v1.6.0 prototype reached 31/35 on a Qwen3-14B math benchmark whose FP16 baseline sits at $\sim$13/35. An uncompressed-model-beating compressed cache is, on its face, impossible; the likely explanations we can think of are (i) an implicit regularization effect from the uniform angular grid acting as a denoiser on a noisy attention distribution, (ii) a mismatch between our scoring rubric and the FP16 baseline on this particular suite, or (iii) a lucky interaction between our alpha-sharpening and Qwen3-14B's specific attention entropy profile. We did not reproduce the effect on Gemma 4 or GPT-OSS. We report the observation as-is rather than prune it from the record: the number happened, and understanding \emph{why} may well be more valuable than the number itself.

\paragraph{Rank vs.\ value as independent axes.} A lesson we draw from the v1.6.0/v1.7.0 arc is that token-rank compression (which tokens to keep) and bit-depth compression (how to store each kept token) are effectively orthogonal. Our early experiments conflated them---trying to fix math-benchmark quality by adding bits to the quantizer---and improvements plateaued. Separating the axes (TBQ/AMX for value, TriAttention for rank) let each be pushed to its own limit without the other contaminating the design.

% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{KV Cache Quantization.} KIVI~\cite{kivi2024} and KVQuant~\cite{kvquant2024} compress KV caches using per-channel or per-token quantization without transforms. Compared to these direct quantization methods, TurboQuant's WHT preprocessing uniformizes the coefficient distribution, enabling optimal Lloyd-Max quantization regardless of the input structure.

\paragraph{QJL.} The Quantized Johnson-Lindenstrauss lemma~\cite{qjl2024} enables 1-bit random projections that preserve inner products. TurboQuant-prod (TBQP) combines $(b{-}1)$-bit Lloyd-Max with 1-bit QJL on the residual, trading attention score variance for better reconstruction MSE.

\paragraph{Token-level sparsification.} TriAttention~\cite{triattention2026} introduced the trig-summary rank statistic we exploit, and StreamingLLM~\cite{streamingllm2023} established that the earliest tokens act as attention sinks and must be protected from any eviction policy. Our contribution relative to both is the CUDA-native integration with a \emph{quantized} K cache: we show that, in the WHT domain, the rank score and the per-element quantization share the same algebraic substrate, so they compose without dequantization.

\paragraph{Flash Attention.} FlashAttention~\cite{dao2022flashattention,dao2023flashattention2} and its variants optimize attention computation through tiling and online softmax. Our implementation extends the llama.cpp \texttt{fattn-vec} kernel, which targets single-token decoding with full warp utilization.

\paragraph{Attention Score Correction.} SmoothQuant~\cite{smoothquant2023} and outlier-aware methods address activation quantization but do not correct the softmax flattening caused by KV quantization noise. Our dynamic sharpening is, to our knowledge, the first principled correction derived from the quantization SQNR.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented the first production implementation of TurboQuant KV cache compression, achieving 4.2$\times$ raw memory reduction with 6\% perplexity cost on a state-of-the-art MoE model, and extending this to an effective $\sim$4.74$\times$ at inference time on Qwen3-14B by composing TurboQuant's value-axis compression with a CUDA-native port of TriAttention along the rank axis. Our dynamic attention sharpening formula provides a theoretically grounded, hyperparameter-free correction for quantization-induced softmax flattening. The discovery of the half-precision amplification pathology highlights that numerical precision requirements for WHT-based methods differ fundamentally from standard quantization---a consideration absent from the original paper that may be relevant to future WHT-based compression research. The TriAttention integration, in turn, demonstrates that Parseval's identity is the bridge between rank-level and value-level cache compression: the same sign-Hadamard rotation that enables Lloyd-Max also leaves the trig-summary score invariant, so the two compositions require no host roundtrip and no dequantization.

The full implementation, supporting 22 quantization variants across five head dimensions, is available at \url{https://github.com/AmesianX/TurboQuant} as a fork of llama.cpp.

% ============================================================
\section*{Acknowledgments}

This work was conducted independently on an NVIDIA DGX Spark. The author thanks the llama.cpp community and the TurboQuant authors at Google DeepMind for the theoretical foundation. The TriAttention integration in v1.7.0 directly builds on the work of Mao et al.~\cite{triattention2026}, and our CUDA port was informed by the open-source Python reference implementation \texttt{domvox/triattention-ggml}; their prior art in structuring the scoring and eviction loop saved weeks of design time.

% ============================================================
\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{turboquant2026}
S.~Ashkboos, A.~Mohtashami, R.~Pramanik, Y.~Li, A.~Jaggi, D.~Alistarh.
\newblock TurboQuant: Online KV Cache Quantization via Hadamard Transform.
\newblock In \emph{Proceedings of ICLR}, 2026.

\bibitem{dao2022flashattention}
T.~Dao, D.~Fu, S.~Ermon, A.~Rudra, C.~R\'{e}.
\newblock FlashAttention: Fast and Memory-Efficient Exact Attention with IO-Awareness.
\newblock In \emph{NeurIPS}, 2022.

\bibitem{dao2023flashattention2}
T.~Dao.
\newblock FlashAttention-2: Faster Attention with Better Parallelism and Work Partitioning.
\newblock In \emph{ICLR}, 2024.

\bibitem{kivi2024}
Z.~Liu, J.~Yuan, H.~Jin, S.~Zhong, Z.~Xu, V.~Braverman, B.~Chen, X.~Hu.
\newblock KIVI: A Tuning-Free Asymmetric 2bit Quantization for KV Cache.
\newblock In \emph{ICML}, 2024.

\bibitem{kvquant2024}
C.~Hooper, S.~Kim, H.~Mohammadzadeh, M.~W.~Mahoney, Y.~S.~Shao, K.~Keutzer, A.~Gholami.
\newblock KVQuant: Towards 10 Million Context Length LLM Inference with KV Cache Quantization.
\newblock \emph{arXiv preprint arXiv:2401.18079}, 2024.

\bibitem{qjl2024}
A.~Zandieh, M.~Daliri, I.~Han.
\newblock QJL: 1-Bit Quantized JL Transform for KV Cache Quantization with Zero Overhead.
\newblock In \emph{NeurIPS}, 2024.

\bibitem{smoothquant2023}
G.~Xiao, J.~Lin, M.~Seznec, H.~Wu, J.~Demouth, S.~Han.
\newblock SmoothQuant: Accurate and Efficient Post-Training Quantization for Large Language Models.
\newblock In \emph{ICML}, 2023.

\bibitem{triattention2026}
Y.~Mao et al.
\newblock TriAttention: Trigonometric Importance Scoring for Training-Free KV Cache Eviction.
\newblock \emph{arXiv preprint arXiv:2604.04921}, 2026.

\bibitem{streamingllm2023}
G.~Xiao, Y.~Tian, B.~Chen, S.~Han, M.~Lewis.
\newblock Efficient Streaming Language Models with Attention Sinks.
\newblock In \emph{ICLR}, 2024.

\end{thebibliography}

\end{document}
